A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2024-02-12 paper/paper.tex
Add more text

\mathit{Contexts}\\
[[G]]       & ::= & [[empty]]\ |\ [[G ++ A]] &  \\ \\
\mathit{Terms}\\
[[a]],[[b]],[[c]],[[t]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[i]]\  |\ [[Void]]
                  & \mbox{universes, variables, empty type} \\
            & |   & [[Pi A B]]\ |\ [[\ A a]]\ |\ [[a b]]
                  & \mbox{function types, abstractions, applications} \\
            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J t a b p]]
                  & \mbox{equality types, reflexivity proof, J eliminator} \\
            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ |\  [[if a b0 b1]]
                  & \mbox{boolean type, true, false, if} \\ \\
\mathit{Renaming}\\
[[xi]] & \in & [[SNat -> SNat]] & \\ \\
\mathit{Substitution}\\
[[rho]] & \in & [[SNat -> STm]] &
\end{array}
\]
  \caption{Syntax of \lang}
  \label{fig:syntax}
\end{figure}

\begin{figure}[h]
    \[
      \begin{array}{lll}
        \mathit{Curried\ Addition} \\
        add(n) & := & m \mapsto n + m \\ \\
        \mathit{Extension ([[xi]])} \\
        [[(xi .: j) 0]]  & := & 0 \\
        [[(xi .: j) Suc i]]  & := & [[xi j]] \\ \\

        \mathit{Extension ([[rho]])} \\
        [[(rho .: a) 0]]  & := & [[a]] \\
        [[(rho .: a) Suc i]]  & := & [[rho i]] \\ \\

        \mathit{Up ([[xi]])} \\
        [[up xi]] & := & (add(1) \circ [[xi]]) , 0 \\ \\

        \mathit{Renaming} \\
        [[i {xi}]] & := & [[xi i]] \\
        [[(Set i) {xi}]] & := & [[Set i]] \\
        [[Void {xi}]] & := & [[Void]]\\
        [[(Pi A B) {xi}]] & := & [[Pi A{xi} B{up xi}]] \\
        [[(\ A a) {xi}]] & := & [[\ A {xi} (#a {up xi}#)]] \\
        [[(a b) {xi}]] & := & [[a {xi} (# b {xi} #)]] \\
        [[Bool {xi}]] & := & [[Bool]] \\
        [[true {xi}]] & := & [[true]] \\
        [[false {xi}]] & := & [[false]] \\
        [[(if a b0 b1) {xi}]] & := & [[if a{xi} b0{xi} b1{xi}]] \\
        [[(a ~ b : A) {xi}]] & := & [[ a{xi} ~ b{xi} : A{xi}]] \\
        [[refl {xi}]] & := & [[refl]] \\
        [[(J t a b p ) {xi}]] & := & [[J t {xi} a {xi} b {xi} p{xi}]] \\ \\

        \mathit{Shift} \\
        [[up n A]] & := & A \langle add(n) \rangle \\ \\

        \mathit{Shift\ one} \\
        [[up1 A]] & := & \uparrow^1 [[A]] \\ \\

        \mathit{Lookup} \\
        [[(G ++ A) 0]] & := &  [[A]] \\
        [[(G ++ A) Suc i]] & := & [[G i]] \\ \\

        \mathit{Drop} \\
        [[drop 0 G]] & := & [[G]] \\
        [[drop Suc i (#G ++ A#)]] & := & [[drop i G]] \\ \\
      \end{array}
    \]
  \caption{Auxiliary Functions over Syntax}
  \label{fig:auxdef}
\end{figure}

\begin{figure}[h]
\drules[P]{$[[a => b]]$}{Parallel Reduction}{Var, Set, Void, Pi, Abs, App, AppAbs, True, False, If, IfTrue, IfFalse, Bool, Eq, Refl, J, JRefl}
\drules[PS]{$[[a =>+ b]]$}{Transitive Closure of Parallel Reduction}{One, Step}
\drules[C]{$[[a <=> b]]$}{Coherence}{Intro}
\caption{Parallel reduction and coherence}
\label{fig:par}
\end{figure}

\begin{figure}[h]
\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{Empty, Cons}
\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv, Refl, J, Bool, True, False, If, Void}
\caption{Syntactic typing for \lang}
\label{fig:typing}
\end{figure}

@@

Figure~\ref{fig:par} shows the definition of the parallel reduction
relation, which takes the form $[[a => b]]$. We use $[[a =>+ b]]$ to
represent its transitive and reflexive closure, which in turn allows us to define
the coherence relation $[[a <=> b]]$.
We say that two terms $[[a]]$
and $[[b]]$ are coherent if they can eventually reduce to some common
term $[[c]]$ through parallel reduction. The symmetric notation of
coherence suggests that it is an equivalence relation, and in fact, we
will use coherence as the equational theory for our type conversion
rule.

We sketch out some key properties about coherence without giving their
proofs. Our technique for establishing those results is based on
\citet{takahashi-parallel-reduction} and a modern exposition of the
same technique can be found in \citet{plfa22.08}.

TODO: remove the lengthy discussion below

Now, we prove that coherence is indeed an equivalence relation.

First, we show that coherence is reflexive through the following
sequence of lemmas.
\begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
  \label{lemma:parrefl}
  For all terms $[[a]]$, $[[a => a]]$.
\end{lemma}

\begin{lemma}[Pars refl]
  \label{lemma:parsrefl}
  For all terms $[[a]]$, $[[a =>+ a]]$.
\end{lemma}

\begin{lemma}[Coherence refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
  \label{lemma:coherencerefl}
  For all terms $[[a]]$, $[[a <=> a]]$.
\end{lemma}

Lemma~\ref{lemma:parrefl} can be proven by structural induction over
the term $[[a]]$. Lemmas~\ref{lemma:parsrefl} and \ref{lemma:coherencerefl}
immediately follow as corollaries of Lemma~\ref{lemma:parrefl}.

The reflexivity of parallel reduction enables us to embed rules from
call-by-name semantics into parallel reduction, as the following lemma
shows.
\begin{lemma}[Par AbsCbn\footnote{\dotv{join.v}{P\_AppAbs\_cbn}}] For all $[[A]], [[a]],$ and $[[b]]$,
  \label{lemma:parabscbn}
  $[[(\ A a) b => b {a}]]$
\end{lemma}
\begin{proof}
  Immediate from Lemma~\ref{lemma:parrefl} and \rref{P-AppAbs}.
\end{proof}

Symmetry of coherence immediately falls from its definition.
\begin{lemma}[Coherence sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
  \label{lemma:coherencesym}
  If $[[a <=> b]]$, then $[[b <=> a]]$.
\end{lemma}

Before we can prove transitivity, we need to show that parallel
reduction satisfies the diamond property.
\begin{lemma}[Par cong\footnote{\dotv{join.v}{par\_cong}}]
  \label{lemma:parcong}
  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 } => a1 { b1 }]]$.
\end{lemma}
\begin{lemma}[Par diamond\footnote{\dotv{join.v}{par\_confluent}}]
  \label{lemma:pardiamond}
  If $[[a => b0]]$ and $[[a => b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 => c]]$ and $[[b1 => c]]$.
\end{lemma}
The congruence property (Lemma~\ref{lemma:parcong}) can be proven by
structural induction over the derivation of $[[a0 => a1]]$.
Likewise, Lemma~\ref{lemma:pardiamond} can be proven by structural induction
over the derivation of $[[a => b0]]$. The \rref{P-AppAbs} case requires
the use of Lemma~\ref{lemma:parcong}.

From Lemma~\ref{lemma:parcong} and \ref{lemma:parrefl}, we recover the
single substitution property as a simple corollary.
\begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
  \label{lemma:parsubst}
  If $[[a0 => a1]]$, then $[[a0 {b} => a1 {b}]]$ for arbitrary $[[b]]$.
\end{corollary}

A relation that satisfies the
diamond property must also be confluent, meaning that its transitive
and reflexive closure is confluent.
\begin{lemma}[Par confluent\footnote{\dotv{join.v}{pars\_confluent}}]
  \label{lemma:parconfluent}
  If $[[a =>+ b0]]$ and $[[a =>+ b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 =>+ c]]$ and $[[b1 =>+ c]]$.
\end{lemma}
While $[[a =>+ b]]$ is defined as the transitive closure of $[[a => b]]$,
it coincides with the transitive and reflexive closure of $[[a => b]]$ since $[[a
=> b]]$ is reflexive (Lemma~\ref{lemma:parrefl}).

The transitivity of the coherence relation follows as a corollary of
Lemma~\ref{lemma:parconfluent}.
\begin{lemma}[Coherence trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}
By the definition of coherence, there exists some term $[[b]]$ such that $[[a0 =>+ b0]]$,
$[[a1 =>+ b0]]$ and some term $[[b1]]$ such that $[[a1 =>+ b1]]$ and
$[[a2 =>+ b1]]$. By Lemma~\ref{lemma:parconfluent}, there exists some
term $[[c]]$ such that $[[b0 =>+ c]]$ and $[[b1 =>+ c]]$. It sufficies
to show that $[[a0 =>+ c]]$ and $[[a2 =>+ c]]$, both of which
trivially hold since the transitive closure $[[a =>+ b]]$ is transitive.
This concludes the proof that coherence is an equivalence relation.
\begin{lemma}[Coherence Equivalence]
  \label{lemma:coherenceequiv}
  The relation $[[a <=> b]]$ satisfies reflexivity, symmetry, and
  transitivity and therefore is an equivalence relation.
\end{lemma}

Figure~\ref{fig:typing} gives the full typing rules for
\lang{}. \Rref{T-Conv} uses the definition of coherence from earlier
for type conversion. The use of an untyped relation for type
conversion makes our formulation slightly different from languages
such as MLTT, where the judgmental equality typically takes the form
TODO, where terms ... and ... are known to be well-typed. The
equivalence of such systems and a system that uses untyped equality
are explored in detail in ...

Without fancy eta laws, it is easy to embed a typed language into an
untyped language.

@@

\begin{figure}[h]
\begin{equation*}
    [[InterpR i A S]] := [[ Interp I i A S  ]], \text{where } [[I i]] := [[{A | exists S , InterpR i A S}]]
\end{equation*}
\caption{Logical relation for all universe levels}
\label{fig:logrelrec}
\end{figure}
To tie the knot and obtain an interpretation of all universe levels,
we define in Figure~\ref{fig:logrelrec} the final version of our interpretation judgment recursively
using the well-foundedness of the strict less than relation on natural
numbers (recall that
the parameter $[[I]]$ of $[[Interp I i A S]]$ takes only natural
numbers strictly less than $[[i]]$ as its input).
The judgment $[[InterpR i A S]]$ now reads that the type $[[A]]$ is a
level-$[[i]]$ type \emph{semantically} inhabited by terms from the set
$[[S]]$. For the
majority of the properties we are about to prove in this section, we
do not need any information about the parameterized function $[[I]]$.
Each property about $[[InterpR i A S]]$ follows as a corollary of
a property about $[[Interp I i A S]]$ with no or few assumptions imposed on
$[[I]]$. As a result, we usually state our lemmas in terms of
$[[Interp I i A S]]$ without duplicating them in terms of $[[InterpR i
A S]]$.

For the rest of the section, we establish some important facts about
the logical relation that will be useful for proving fundamental theorem
in Section~\ref{sec:logrelproof}.

The relation $[[Interp I i A S]]$ satisfies the following inversion
principles.
\begin{lemma}[Inversion of the logical relation]
  \label{lemma:interpinv}\leavevmode
  \begin{enumerate}
  \item If $[[Interp I i Void S]]$, then $[[S = emptyset]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Void\_inv}}
  \item If $[[Interp I i Bool S]]$, then $[[S = { a | a =>+ true \/ a =>+ false   }]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Bool\_inv}}
  \item If $[[Interp I i a ~ b : A S]]$, then $[[S = { p | p =>+ refl , a <=> b  }]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Eq\_inv}}
  \item If $[[Interp I i Pi A B S0]]$, then there exists $[[S]],[[F]]$ such that:\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_inv}}
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
  \item If $[[Interp I i Set j S]]$, then $[[j < i]]$ and $[[S = I j]]$.\footnote{\dotv{semtyping.v}{InterpExt\_Univ\_inv}}
  \end{enumerate}
\end{lemma}
\begin{proof}
  We only show the most involved function case; the rest follows a
  similar but simpler pattern. We start by
  inducting over the derivation of $[[Interp I i Pi A B S]]$. There
  are only two possible cases we need to consider.
  \begin{description}
  \item[\Rref{I-Pi}:] Immediate.
  \item[\Rref{I-Red}:] We are given that $[[Interp I i Pi A B S0]]$.
    We know that there exists some $[[A0]]$ and
    $[[B0]]$ such that $[[Pi A B => Pi A0 B0]]$ and $[[Interp I i Pi
    A0 B0 S0]]$. From the
    induction hypothesis, there exists $[[S]]$ and $[[F]]$ such that :
    \begin{itemize}
    \item $[[Interp I i A0 S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B0 { a } F a #)]]$
    \item $[[S0 = { b | forall a, (# a in S implies b a in F a #) }]]$
    \end{itemize}
    By inverting $[[Pi A B => Pi A0 B0]]$, we derive $[[A => A0]]$ and
    $[[B => B0]]$. By Lemma~\ref{lemma:parsubst}, we have $[[B {a} => B0 {a} ]]$ for all
    $[[a]]$. As a result, by \rref{I-Red}, the same $[[S]]$ and
    $[[F]]$ satisfies the following properties:
    \begin{itemize}
    \item $[[Interp I i A S ]]$
    \item $[[F in S -> PowerSet STm]]$
    \item $[[forall a, (# a in S implies Interp I i B { a } F a #)]]$
    \end{itemize}
    The properties above are exactly the preconditions needed to apply
    \rref{I-Pi} for $[[Pi A B]]$ to finish off the proof.
  \end{description}
\end{proof}

>>>>

\mathit{Contexts}\\
[[G]]       & ::= & [[empty]]\ |\ [[G ++ A]] &  \\ \\
\mathit{Terms}\\
[[a]],[[b]],[[c]],[[t]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[i]]\  |\ [[Void]]
                  & \mbox{universes, variables, empty type} \\
            & |   & [[Pi A B]]\ |\ [[\ A a]]\ |\ [[a b]]
                  & \mbox{function types, abstractions, applications} \\
            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J t a b p]]
                  & \mbox{equality types, reflexivity proof, J eliminator} \\
            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ % |\  [[if a b0 b1]]
                  & \mbox{boolean type, true, false} \\
            & |   & [[if a b0 b1]]
                  & \mbox{if} \\ \\
\mathit{Renaming}\\
[[xi]] & \in & [[SNat -> SNat]] & \\ \\
\mathit{Substitution}\\
[[rho]] & \in & [[SNat -> STm]] &
\end{array}
\]
  \caption{Syntax of \lang}
  \label{fig:syntax}
\end{figure}

% \begin{figure}[h]
%     \[
%       \begin{array}{lll}
%         \mathit{Curried\ Addition} \\
%         add(n) & := & m \mapsto n + m \\ \\
%         \mathit{Extension ([[xi]])} \\
%         [[(xi .: j) 0]]  & := & 0 \\
%         [[(xi .: j) Suc i]]  & := & [[xi j]] \\ \\

%         \mathit{Extension ([[rho]])} \\
%         [[(rho .: a) 0]]  & := & [[a]] \\
%         [[(rho .: a) Suc i]]  & := & [[rho i]] \\ \\

%         \mathit{Up ([[xi]])} \\
%         [[up xi]] & := & (add(1) \circ [[xi]]) , 0 \\ \\

%         \mathit{Renaming} \\
%         [[i {xi}]] & := & [[xi i]] \\
%         [[(Set i) {xi}]] & := & [[Set i]] \\
%         [[Void {xi}]] & := & [[Void]]\\
%         [[(Pi A B) {xi}]] & := & [[Pi A{xi} B{up xi}]] \\
%         [[(\ A a) {xi}]] & := & [[\ A {xi} (#a {up xi}#)]] \\
%         [[(a b) {xi}]] & := & [[a {xi} (# b {xi} #)]] \\
%         [[Bool {xi}]] & := & [[Bool]] \\
%         [[true {xi}]] & := & [[true]] \\
%         [[false {xi}]] & := & [[false]] \\
%         [[(if a b0 b1) {xi}]] & := & [[if a{xi} b0{xi} b1{xi}]] \\
%         [[(a ~ b : A) {xi}]] & := & [[ a{xi} ~ b{xi} : A{xi}]] \\
%         [[refl {xi}]] & := & [[refl]] \\
%         [[(J t a b p ) {xi}]] & := & [[J t {xi} a {xi} b {xi} p{xi}]] \\ \\

%         \mathit{Shift} \\
%         [[up n A]] & := & A \langle add(n) \rangle \\ \\

%         \mathit{Shift\ one} \\
%         [[up1 A]] & := & \uparrow^1 [[A]] \\ \\

%         \mathit{Lookup} \\
%         [[(G ++ A) 0]] & := &  [[A]] \\
%         [[(G ++ A) Suc i]] & := & [[G i]] \\ \\

%         \mathit{Drop} \\
%         [[drop 0 G]] & := & [[G]] \\
%         [[drop Suc i (#G ++ A#)]] & := & [[drop i G]] \\ \\
%       \end{array}
%     \]
%   \caption{Auxiliary Functions over Syntax}
%   \label{fig:auxdef}
% \end{figure}

@@

\begin{figure}[h]
\begin{minipage}{0.9\textwidth}
\drules[P]{$[[a => b]]$}{Parallel Reduction}{Var, Set, Void, Pi, Abs, App, AppAbs, True, False, If, IfTrue, IfFalse,
  Bool, Eq, Refl, J, JRefl}
\drules[PS]{$[[a =>+ b]]$}{Transitive Closure of Parallel Reduction}{One, Step}
\drules[C]{$[[a <=> b]]$}{Coherence}{Intro}
\end{minipage}
\caption{Parallel reduction and coherence}
\label{fig:par}
\end{figure}

Figure~\ref{fig:par} shows the definition of the parallel reduction
relation, which takes the form $[[a => b]]$. We use $[[a =>+ b]]$ to
represent its transitive and reflexive closure, which in turn allows us to define
the coherence relation $[[a <=> b]]$.
We say that two terms $[[a]]$
and $[[b]]$ are coherent if they can eventually reduce to some common
term $[[c]]$ through parallel reduction. The symmetric notation of
coherence suggests that it is an equivalence relation.% , and in fact, we
% will use coherence as the equational theory for our type conversion
% rule.

We sketch out some key properties about parallel reduction and
coherence without giving their proofs. Our technique for establishing
those results is based on \citet{takahashi-parallel-reduction} and a
modern exposition of the same technique can be found in
\citet{plfa22.08}.
\begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
  \label{lemma:parrefl}
  For all terms $[[a]]$, $[[a => a]]$.
\end{lemma}
\begin{lemma}[Par cong\footnote{\dotv{join.v}{par\_cong}}]
  \label{lemma:parcong}
  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 } => a1 { b1 }]]$.
\end{lemma}
\begin{lemma}[Par diamond\footnote{\dotv{join.v}{par\_confluent}}]
  \label{lemma:pardiamond}
  If $[[a => b0]]$ and $[[a => b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 => c]]$ and $[[b1 => c]]$.
\end{lemma}
\begin{lemma}[Coherence refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
  \label{lemma:coherencerefl}
  For all terms $[[a]]$, $[[a <=> a]]$.
\end{lemma}
\begin{lemma}[Coherence sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
  \label{lemma:coherencesym}
  If $[[a <=> b]]$, then $[[b <=> a]]$.
\end{lemma}
\begin{lemma}[Coherence trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}

From Lemma~\ref{lemma:coherencerefl}, \ref{lemma:coherencesym}, and
\ref{lemma:coherencetrans}, we conclude that coherence is indeed an
equivalence relation. It is then trivial to derive that coherence is
exactly the same as the untyped $\beta$-equivalence.

% TODO: remove the lengthy discussion below

% Now, we prove that coherence is indeed an equivalence relation.

% First, we show that coherence is reflexive through the following
% sequence of lemmas.

% Lemma~\ref{lemma:parrefl} can be proven by structural induction over
% the term $[[a]]$. Lemmas~\ref{lemma:parsrefl} and \ref{lemma:coherencerefl}
% immediately follow as corollaries of Lemma~\ref{lemma:parrefl}.

% The reflexivity of parallel reduction enables us to embed rules from
% call-by-name semantics into parallel reduction, as the following lemma
% shows.
% \begin{lemma}[Par AbsCbn\footnote{\dotv{join.v}{P\_AppAbs\_cbn}}] For all $[[A]], [[a]],$ and $[[b]]$,
%   \label{lemma:parabscbn}
%   $[[(\ A a) b => b {a}]]$
% \end{lemma}
% \begin{proof}
%   Immediate from Lemma~\ref{lemma:parrefl} and \rref{P-AppAbs}.
% \end{proof}

% Symmetry of coherence immediately falls from its definition.

% Before we can prove transitivity, we need to show that parallel
% reduction satisfies the diamond property.

% The congruence property (Lemma~\ref{lemma:parcong}) can be proven by
% structural induction over the derivation of $[[a0 => a1]]$.
% Likewise, Lemma~\ref{lemma:pardiamond} can be proven by structural induction
% over the derivation of $[[a => b0]]$. The \rref{P-AppAbs} case requires
% the use of Lemma~\ref{lemma:parcong}.

% From Lemma~\ref{lemma:parcong} and \ref{lemma:parrefl}, we recover the
% single substitution property as a simple corollary.
% \begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
%   \label{lemma:parsubst}
%   If $[[a0 => a1]]$, then $[[a0 {b} => a1 {b}]]$ for arbitrary $[[b]]$.
% \end{corollary}

% A relation that satisfies the
% diamond property must also be confluent, meaning that its transitive
% and reflexive closure is confluent.
% \begin{lemma}[Par confluent\footnote{\dotv{join.v}{pars\_confluent}}]
%   \label{lemma:parconfluent}
%   If $[[a =>+ b0]]$ and $[[a =>+ b1]]$, then there exists some term
%   $[[c]]$ such that $[[b0 =>+ c]]$ and $[[b1 =>+ c]]$.
% \end{lemma}
% While $[[a =>+ b]]$ is defined as the transitive closure of $[[a => b]]$,
% it coincides with the transitive and reflexive closure of $[[a => b]]$ since $[[a => b]]$ is reflexive (Lemma~\ref{lemma:parrefl}).

% The transitivity of the coherence relation follows as a corollary of
% Lemma~\ref{lemma:parconfluent}.
% By the definition of coherence, there exists some term $[[b]]$ such that $[[a0 =>+ b0]]$,
% $[[a1 =>+ b0]]$ and some term $[[b1]]$ such that $[[a1 =>+ b1]]$ and
% $[[a2 =>+ b1]]$. By Lemma~\ref{lemma:parconfluent}, there exists some
% term $[[c]]$ such that $[[b0 =>+ c]]$ and $[[b1 =>+ c]]$. It sufficies
% to show that $[[a0 =>+ c]]$ and $[[a2 =>+ c]]$, both of which
% trivially hold since the transitive closure $[[a =>+ b]]$ is transitive.
% This concludes the proof that coherence is an equivalence relation.
% \begin{lemma}[Coherence Equivalence]
%   \label{lemma:coherenceequiv}
%   The relation $[[a <=> b]]$ satisfies reflexivity, symmetry, and
%   transitivity and therefore is an equivalence relation.
% \end{lemma}

\begin{figure}[ht]
\begin{minipage}{0.9\textwidth}
\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{Empty, Cons}
\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv, Refl, J, Bool, True, False, If, Void}
\end{minipage}
\caption{Syntactic typing for \lang}
\label{fig:typing}
\end{figure}

Figure~\ref{fig:typing} gives the full typing rules for
\lang{}. \Rref{T-Conv} uses the definition of coherence from earlier
as our equality judgment for type conversion. The use of an untyped
relation for type conversion makes our formulation slightly different
from languages such as MLTT, where the judgmental equality typically
takes the form $\Gamma \vdash a \equiv b : A$, which implies $[[G |- a :
A]]$ and $[[G |- b : A]]$, the well-typedness of $[[a]]$ and $[[b]]$.
\citet{siles2012pure} shows the equivalence of Barendregt's
Pure Type System~\citep{barendregt1991introduction}, which employs
untyped equality, and its variant that uses typed judgmental
equality. Furthermore, it is trivial to embed a system with judgmental
equality to a system with untyped equality by erasing typing
information. As a result, it is easy to port our consistency result from our
type system to a variant with typed judgmental equality
as long as the typed system does not include $\eta$ laws
that would require type annotations (e.g. the $\eta$-law for unit
types) and we discuss in Section~\ref{sec:discuss} how we can extend
our proof to handle those rules.

Working with a system with untyped equality has the huge benefit that
the confluence result for untyped parallel reduction
(Lemma~\ref{lemma:pardiamond}) is easily derivable without having to
resort to the complex syntatic (resp. semantic) technique from
\citet{siles2012pure} (resp. \citet{decagda}) to resolve
the circularity of subject reduction and $\Pi$-injectivity described
in \citet{siles2012pure}. Section~\ref{sec:discuss} explains how we
can generalize our technique to include $\eta$-law for functions and
show the existence of normal form for well-typed (open and closed)
terms, achieving a similar level of expressiveness of the type system
and strength of metatheoretic property as \citet{decagda}.

Finally, since our system has an infinite universe hierarchy, we can
present the system à la Russell by using the same judgment form $[[G
|- a : A]]$ regardless of whether $[[a]]$ is a term or a type. There
is therefore no need to distinguish between big types and small types
and duplicate our typing specification.

% , the
% statement that $\Gamma \vdash [[Pi A0 B0]] \equiv [[Pi A1 B1]]$
% implies $\Gamma \vdash [[A0]] \equiv [[A1]]$ and
% $\Gamma, [[A0]]\vdash [[B0]] \equiv [[B1]]$.

% working with a system with untyped equality not only preserves the
% same level of generality,
% Of course,
% but also enables us to derive confluence
% (Lemma~\ref{lemma:parconfluent}) early on without having to use the
% intricate techniques from \citet{lemma:}

% TODO, where terms ... and ... are known to be
% well-typed. The equivalence of such systems and a system that uses
% untyped equality are explored in detail in ...

% Without fancy eta laws, it is easy to embed a typed language into an
% untyped language.

@@

\begin{figure}[h]
\begin{equation*}
    [[InterpR i A S]] := [[ Interp I i A S  ]], \text{where } [[I i]] := [[{A | exists S , InterpR i A S}]]
\end{equation*}
\caption{Logical relation for all universe levels}
\label{fig:logrelrec}
\end{figure}
To tie the knot and obtain an interpretation of all universe levels,
we define in Figure~\ref{fig:logrelrec} the final version of our interpretation judgment recursively
using the well-foundedness of the strict less than relation on natural
numbers (recall that
the parameter $[[I]]$ of $[[Interp I i A S]]$ takes only natural
numbers strictly less than $[[i]]$ as its input).
The judgment $[[InterpR i A S]]$ now reads that the type $[[A]]$ is a
level-$[[i]]$ type \emph{semantically} inhabited by terms from the set
$[[S]]$.

By unfolding the definition from Figure~\ref{fig:logrelrec}, we can
now show that the same introduction rules for $[[Interp I i A S]]$ is
admissible for $[[InterpR i A S]]$. We give as example TODO:

In most informal presentations, instead of defining the logical
relation in two steps as we have shown above, the rules for $[[InterpR
i A S]]$ is given directly, with the implicit understanding of
inductive definitions nested inside a recursive function. We choose
the more explicit definition not only because it is directly definable
in existing proof assistants where inductive definitions must appear
at the top level, but also because it makes clear the induction
principle we are allowed to use when reasoning about $[[InterpR i A
S]]$.

The most general format of the induction principle over $[[InterpR i A
S]]$ is first by strong induction over the universe level $[[i]]$
followed by structural induction over $[[Interp I i A S]]$. As
examples, ... (\rref{I-Set} and \rref{I-Red}).

For the majority of the properties we are about to prove in this section, we
do not need any information about the parameterized function $[[I]]$.
Each property about $[[InterpR i A S]]$ follows as a corollary of
a property about $[[Interp I i A S]]$ with no or few assumptions imposed on
$[[I]]$. As a result, we usually state our lemmas in terms of
$[[Interp I i A S]]$ without duplicating them in terms of $[[InterpR i
A S]]$.

To derive consistency, it suffices to restrict $[[S]]$ to the set of
terms to reduce to closed terms. \Rref{I-Void, I-Bool} are
unsurprising; when considering only closed terms, the empty type
should not be inhabited and therefore corresponds to the empty set,
whereas the boolean type is semantically inhabited by terms that
evaluate to the boolean values $[[true]]$ or $[[false]]$. \Rref{I-Eq}
says that an equality type $[[a ~ b : A]]$ corresonds to the
set of terms that evaluate to $[[refl]]$ when $[[a <=> b]]$ holds and
otherwise corresponds to the empty set. Side conditions like $[[a <=>
b]]$ are typically required for indexed types, of which equality types
are an instance.

\Rref{I-Pi} is the most interesting. The precondition consists of a
mysterious function $[[F]]$ that takes an element from the set $[[S]]$, the interpretation
of the type $[[A]]$, and returns a subset of terms. $[[F]]$ is further
subject to the constraint that $[[forall a, (# a in S implies Interp I
i B { a } F a #)]]$. Ignoring the content below the horizontal bar of
the derivation, the statements from the precondition are equivalent to
the conjunction of the following statements:
\begin{itemize}
\item $[[Interp I i A S]]$
\item $\forall [[a]], \text{ if }[[a in S]]\text{, then } \exists
  [[S0]]\text{, } [[Interp I i B { a } S0 ]] $
\end{itemize}
In general, let $R$ be a binary relation over the sets $[[A]]$ and
$[[B]]$, it easy to verify that the following equivalence:
\[\forall a \in A, \exists b \in B \text{ such that } (a,b) \in R
  \iff \exists F \in A \rightarrow B \text, \forall a \in A, (a, F(a))
  \in R\]

The left-hand side is equivalent to the right-hand side, but has no
mentioning of any functions. In the context of \rref{I-Pi}, through
the equivalence, we can more easily tell what's happening: $[[Pi A
B]]$ has a semantic interpretation of $[[A]]$ can be semantically
interpreted as the set $[[S]]$, and for all terms $[[a in S]]$, the
type $[[B {a}]]$, obtained by substituing $[[a]]$ into the output type
$[[B]]$, has a semantic interpretation.

If we look at $[[Interp I i Pi A B { b | forall a, (# a in S implies b
  a in F a #) }]]$, the conclusion of \rref{I-Pi}, we see why the
presentation with an explicit $[[F in S -> PowerSet STm]]$ is
useful. We want to state the fact that $[[B {a}]]$ has a semantic
interpretation of some set $[[S0]]$, but we also want to refer to the
set $[[S0]]$ when we talk about the set that $[[Pi A B]]$ corresponds
to. If we used the alternative representation of the preconditions, we
would be forced to define our logical relation in the following way.

TODO: put the wrong logical relation that violates ...

Unfortunately, both versions not only violate the syntactic strict
positivity constraint required in proof assistants, but
are in fact genuinely not monotone over the domain of
relations.
Intuitively, the failure of monotonicity stems from the fact
that the witness picked in the precondition is not necessarily the
same witness being referred to in the post condition. While it might
be possible to restrict the domain with additional constraints such as
functionality and inversion properties, we opt for our current
skolemized formulation of \rref{I-Pi} so we immediately obtain a
well-defined inductive relation and a usable induction principle. The
more intuitive introduction rules can be later shown to be admissible
a posteriori.

In the rest of the section, we establish some important facts about
the logical relation that will be useful for proving fundamental theorem
in Section~\ref{sec:logrelproof}.
